Depression and suicidality as costly signals of need

### 2021-04-11 Paper.rmd
Minor edits throughout the paper

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might be the result of adaptations for reliably signaling need and bargaining for increased support from reluctant others. To test this hypothesis, we examined responses to  emotional signals in a vignette study in which the existence of conflicts of interest and private information about one's need were emphasized in four scenarios involving claims of extreme need. Although the effect of signal type varied by vignette, costlier signals like depression and suicide attempts resulted in greater perceptions of need and increased likelihood of support compared to simple verbal requests and crying without further symptoms in our sample of 1,240 participants recruited from Amazon Mechanical Turk. Additionally, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on believe in need as predicted. Our results suggest that depression and suicide attempts, apparent human universals, are a relatively successful way of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

One possible reason for the connection between depression, suicide, and social adversity is that some instances of depression and suicide attempts serve as means to elicit increased support from others [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], particularly in times when an individual cannot solve substantial fitness-relevant challenges on their own [@Hagen2003; @SymeEtAl2016]. According to the bargaining models of depression and suicide, the idea is that the small groups humans lived in for much of our evolutionary history created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop [@Hagen2003; @SymeEtAl2016]. In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others valued an individual's cooperation sufficiently and believed them to be in genuine need.

@@

@@

Despite being looked for less frequently, beneficial responses to depressed individuals are also reported. This is seen in that depression has been associated with reduced likelihood of aggression within families [@hops1987; @dadds1992], including those already experiencing high levels of conflict [@SheeberEtAl2001]. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition [@stephens1987].

Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for them resulting in beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, even though suicide was rarely if ever the main topic [@SymeEtAl2016].

# Study aims and predictions

The primary aim of this study was to examine responses to various forms of emotional signaling in an experimental vignette study. We predicted that costlier signals would increase the participants' self-reported belief in the signaler's claims and the likelihood of them providing help, with this prediction entailing the following rank from low to high cost: verbal request, crying, mild-depression, depression, and suicide attempt. We also predicted that changes in the likelihood participants would provide support would be mediated by changes in the participants' belief in the signaler's need pre-signal to post-signal.

@@

As many depression scales include measures of crying [for discussion on the relationship between crying and depression see @bylsma2020] and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012], each signal was designed to build off of the less costly signals so that each signal contained elements of any signal it was costlier than. Although referred to as signals throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, will also provide information to others in the form of cues [For discussion on the evolution of cues from signals see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The full text for the signals are available in the SI.

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true (T3Action). In the US sample, this involved video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. As a final attention check, we presented participants with the T3Action question but included instructions to ignore this and drag the slider all the way to the left (T4AttentionCheck). We did not exclude anyone from the study for failing this nor did we withhold payment due to the time participants had invested in the study at this point.

@@

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (i.e., the assaults or incompatibility with a potential husband in the context of arranged marriage) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in ways that would still be consistent with signaling theory.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as belief in need of financial support in the pilot study. However, the incomplete mediation in this study may have arisen if participants viewed the signal as providing information about adveristy outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

@@

The main limitation of this study is its lack of ecological validity, something which had to be sacrificed for any experimental design involving depression and suicide attempts. It is almost certain that participants experienced this study differently than they would if the event actually happened to them. This may have biased results in a pro-signaler direction if the lack of meaningful costs made support feel less costly or if the likely inability of participants to imagine themselves in an anonymous scenario [@hagen2006] led them to take an opportunity to signal generosity cheaply. It may have also biased participants against helping if they could not fully imagine the characters in the story as kin or important cooperative partners, while the survey's short duration may have weakened the strength of the costlier signals as bargaining tools. However, the fact that the relationships among signal cost, belief in need, and likelihood of providing support all largely fit existing theory (narrow and broad) indicates that participants paid attention to the relevant details and responded in ways that suggest some degree of external validity (needs improvement. For this reason, the benefits of addressing causal relationships that came through our experimental vignette approach likely outweighed the costs.

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. Although possible benefits to depressive states are rarely examined, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can be simultaneously aversive and elicit support [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the responses of others to depressed individuals, especially from those the depressed are in conflict with and those with the capability to provide meaningful benefits.

>>>>

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might be the result of adaptations for reliably signaling need and bargaining for increased support from reluctant others. To test this hypothesis, we examined responses to  emotional signals in a vignette study in which the existence of conflicts of interest and private information about one's need were emphasized in four scenarios involving claims of substantial need. Although the effect of signal type varied by vignette, costlier signals like depression and suicide attempts resulted in greater perceptions of need and increased likelihood of support compared to simple verbal requests and crying without further symptoms in our sample of 1,240 participants recruited from Amazon Mechanical Turk. Additionally, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on believe in need as predicted. Our results suggest that depression and suicide attempts, apparent human universals, are a relatively successful way of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

One possible reason for the connection between depression, suicide, and social adversity is that some instances of depression and suicide attempts serve as means to elicit increased support from others [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], particularly in times when an individual cannot solve substantial fitness-relevant challenges on their own [@Hagen2003; @SymeEtAl2016]. According to the bargaining models of depression and suicide, the small groups humans lived in for much of our evolutionary history created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop [@Hagen2003; @SymeEtAl2016]. In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others valued an individual's cooperation sufficiently and believed them to be in genuine need.

@@

<!-- some overalp with my wikipedia article -->

@@

Despite being looked for less frequently, beneficial responses to depressed individuals are also reported. Depression has been associated with reduced likelihood of aggression within families [@hops1987; @dadds1992], including those already experiencing high levels of conflict [@SheeberEtAl2001]. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition [@stephens1987].

Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, even though suicide was rarely if ever the main topic [@SymeEtAl2016].

## Study aims and predictions

The primary aim of this study was to examine responses to various forms of emotional signaling in an experimental vignette study. As costlier signals are expected in increase the reliability of the information provided, we predicted that costlier signals would increase self-reported belief in the signaler's claims and the likelihood of participants providing help, with this prediction entailing the following rank from low to high cost: verbal request, crying, mild-depression, depression, and suicide attempt. We also predicted that changes in the likelihood participants would provide support would be largely mediated by changes in the participants' belief in the signaler's need pre-signal to post-signal as expected with a costly signaling framework.

@@

As many depression scales include measures of crying [for discussion on the relationship between crying and depression see @bylsma2020] and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012], each signal was designed to build off of the less costly signals so that each signal contained elements of any signal it was costlier than. Although referred to as signals throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, will also provide information to others in the form of cues [for discussion on the evolution of cues from signals see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The full text for the signals are available in the SI.

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true (T3Action). In the US sample, this involved video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. As a final attention check, we presented participants with the T3Action question but included instructions to ignore this and drag the slider all the way to the left (T4AttentionCheck). We did not exclude anyone from the study for failing this nor did we withhold payment.

@@

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (i.e., the assaults or incompatibility with a potential husband in the context of arranged marriage) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in addition to information relevant to whether the event claimed actually happened.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as belief in need of financial support in the pilot study when comparing both depression and suicide attempts (confirm suicide) to verbal requests. However, the partial mediation in this study may have also arisen if participants viewed the signal as providing information about adveristy outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

@@

The main limitation of this study is its lack of ecological validity, something which had to be sacrificed for any experimental design involving depression and suicide attempts. It is almost certain that participants experienced this study differently than they would if the event actually happened to them. This may have biased results in a pro-signaler direction if the lack of meaningful costs made support feel less costly or if the likely inability of participants to imagine themselves in an anonymous scenario [@hagen2006] led them to take an opportunity to signal generosity cheaply. It may have also biased participants against helping if they could not fully imagine the characters in the story as kin or important cooperative partners, while the survey's short duration may have weakened the strength of the costlier signals as bargaining tools. However, the fact that the relationships among signal cost, belief in need, and likelihood of providing support all largely fit existing theory (narrow and broad) indicates that participants paid attention to the relevant details and responded in ways that suggest some degree of external validity (needs improvement). For this reason, the benefits of addressing causal relationships that came through our experimental vignette approach likely outweighed the costs.

@@

The need to explain suicide's connection to depression is often overlooked by explanations for major depression, particularly those which propose non-signaling functions. One explanation for their connection, is that both suicidality and depression may function to limit negative fitness outcomes in times of substantial adversity [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], with costlier signals being required to elicit support when the payoffs to faking need are high enough so that less costly signals are not reliable indicators of one's need [@Rosenthal1993]. For this hypothesis to be correct, instrumental support would need to be provided frequently enough with large enough impacts so that the costs of signaling are outweighed by the benefits on average. In support of this, we found exposure to depression and suicide attempts resulted in greater belief in need and increased likelihood of support in fictional scenarios. This coupled with findings that depression can simultaneously elicit both positive and negative responses [@HokansonEtAl1986; @stephens1987], suggests that more attention should be paid to the responses of others to depressed individuals, especially from those the depressed are in conflict with and those with the capability to provide meaningful benefits.
